Joint maximum-likelihood of phylogenies and ancestral states is not consistent

### 2018-06-28 joint_inf.tex
removing/rewriting to update results in main text

\begin{abstract}
Maximum likelihood estimation in phylogenetics requires a means of handling unknown ancestral states.
Classical maximum likelihood averages over these unknown intermediate states, leading to consistent estimation of the topology and continuous model parameters.
Recently, a computationally-efficient approach was proposed to jointly maximize over these unknown states and phylogenetic parameters.
Although this method of joint maximum likelihood estimation can obtain estimates more quickly, its properties as an estimator are not yet clear.
In this paper, we show that this method of jointly estimating phylogenetic parameters along with ancestral states is not consistent in general.
We find a set of parameters that generate data under a four-taxon tree for which this joint method estimates an incorrect topology in the limit of infinite-length sequences.
For branch length estimation on the correct topology, we outline similar cases where branch length estimates are consistently and heavily biased.
\end{abstract}

@@

In this paper, we show that the joint inference of trees and ancestral sequences is not consistent in general.
To do so, we use a binary symmetric model with data being generated on the four-taxon ``inverse Felsenstein'' tree (a.k.a.\ the ``Farris'' tree) and we construct bounds on the joint objective function to demarcate a sizeable area of long branch lengths in which joint inference is guaranteed to give the wrong tree in the case of perfect sequence data with an infinite number of sites.
We find similar areas where joint inference consistently overestimates interior branch lengths when the topology is known and fixed.

@@

The alternative approach \citep{Sagulenko2018-xl} does away with the marginalization and directly estimates the maximum likelihood parameters of the fully-observed likelihood in \eqref{eq:full_likelihood}.
This is known in statistics as a profile likelihood \citep{Murphy2000-ry}, which exists here because $\alphabet$ is a finite set:
\begin{equation}
\label{eq:profile_likelihood}
L_\nCols'(\tau, t; \fullAlignment) = \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i\in\alphabet^{\nAncestralStateRows}} \ \Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) = \max_{\fullAncestralStates\in\alphabet^{\nAncestralStateRows\times\nCols}} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates).
\end{equation}
We use $\hat{\fullAncestralStates}$ to denote an estimate for $\fullAncestralStates$ obtained by maximizing \eqref{eq:profile_likelihood}, and estimate a topology and branch fidelities using this profile likelihood as
\begin{equation}
\label{eq:profile_likelihood_topology_bl}
(\hat{\tau}, \hat{t}) = \argmax_{\tau, t} \ L_\nCols'(\tau, t; \fullAlignment).
\end{equation}
In general, the functional form of \eqref{eq:profile_likelihood} is determined by inequalities arising from taking maxima row-wise in Tables~\ref{tab:farris_likelihoods} and~\ref{tab:fels_likelihoods} to obtain the likelihood term for each most likely ancestral state, these terms depending on the unknown $(\tau, t)$.
For this reason, in practice, the joint inference strategy estimates $\hat{\fullAncestralStates}$ for a fixed $(\tau,t)$, then $(\hat{\tau},\hat{t})$ given $\hat{\fullAncestralStates}$, maximizing each of these conditional objectives until convergence \citep{Sagulenko2018-xl}.

@@

\subsection*{Inconsistency in topology estimation}

To show an inconsistency in topology estimation, we start with true generating parameters $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ on the InvFels topology (Fig.~\ref{fig:farris-fels-top}a).
We show that, as $\nCols\rightarrow\infty$, there exist values for $x^*$ and $y^*$ such that the value of the likelihood after maximizing using joint inference is greater for the Felsenstein topology ($\tau_2$) than for the true, generating InvFels topology ($\tau_1$).
To do so, we construct an upper bound $C_0(x^*, y^*)$ for the likelihood given the InvFels topology as a function of $x^*$ and $y^*$ and, similarly, a lower bound $C_1(x^*, y^*)$ for the likelihood given the Felsenstein topology.
When $C_0(x^*, y^*) < C_1(x^*, y^*)$, the likelihood in the Felsenstein case is larger than the likelihood in the InvFels case, demonstrating inconsistency (Fig.~\ref{fig:inconsistency-farris}).
\begin{restatable}{theorem}{topoInconsist}
Let $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ and $t=\{x, y, x, y, w\}$.
There exist $C_0(x^*, y^*), C_1(x^*, y^*),$ and a set of $0 < x^*, y^* < 1$ such that
\[
\max_{t} \ \ell_{\tau_1,t^*}(\tau_1, t) \le C_0(x^*, y^*),
\]
\[
C_1(x^*, y^*) \le \max_{t} \ \ell_{\tau_1,t^*}(\tau_2, t)
\]
with $C_0(x^*, y^*) < C_1(x^*, y^*)$.
\end{restatable}
The proof of this theorem is by a detailed examination of inequalities.
Intuitively, $\tau_2$ is favored in performing joint inference since the objective function for $\tau_2$ has more ``degrees of freedom''---Table~\ref{tab:likelihoods} shows that $\tau_1$ has only three possible forms for its objective function while $\tau_2$ has many more.
This enables more possible maxima for $\tau_2$ even if data are generated from $\tau_1$.

\begin{figure}
\centering
% obtained by running
% python joint_inf_plot.py --analytic --topology --delta .001 --plot-name figures/topology-inconsistency.svg
\includegraphics[width=\textwidth]{topology-inconsistency-inkscape}
\caption{
    An analytically-derived region of topological inconsistency in terms of probabilities of a character change along a branch for ``perfect'' data generated on the InvFels topology (Fig.~\ref{fig:farris-fels-top}) with $p_{w^*} = p_{y^*}$ (in terms of fidelities, $w^*=y^*$).
    Due to the looseness of the upper and lower bounds, the parameters in the white region do not necessarily indicate consistency, though all parameters in the shaded region result in an inconsistency.
}
\label{fig:inconsistency-farris}
\end{figure}

\subsection*{Inconsistency in branch length estimation}

We now consider the problem of branch length estimation on the correct tree using joint estimation.
As described above, we use the notion of fidelities to analyze what happens as changes occur along each branch as opposed to the standard branch length formulation.
We analyze two settings on the InvFels tree, corresponding to whether some branch fidelities are fixed at their true values or not.
As above, assume that data is generated from the InvFels tree with two top branches of fidelity $x^*$ and all other branches of fidelity $y^*$ (Fig.~\ref{fig:farris-fels-top}).

In the general case, we do the same but with
\begin{equation}
\label{eq:general-bl-opt}
(\hat{x}, \hat{y}, \hat{w}) = \arg\max_{x,y,w} \ \ell_{\tau_1,t^*}(\tau_1, \{x,y,x,y,w\})
\end{equation}
to find a region where the inferred values do not converge to the generating values.
In the Appendix we consider a ``restricted'' case where we show that, for a nontrivial subset of possible values for $x^*$ and $y^*$, the interior branch fidelity parameter $w$ will be consistently overestimated as exactly equal to one (zero branch length) instead of its true value of $y^*$.
That is, if we estimate $x$ and $y$ correctly, then, for
\[
\hat{w} = \arg\max_{w} \ \ell_{\tau_1,t^*}(\tau_1, \{x^*,y^*,x^*,y^*,w\}),
\]
there is a set of values for $x^*$ and $y^*$ where $\hat{w}\equiv 1$.
These situations are in contrast to the approach using marginal likelihood where $\hat{w}$ necessarily converges to $y^*$ as the number of observations grows \citep{RoyChoudhury2015-ta}.

%Looking to the previous section, the region where $\hat{w}=1$ will still be given by the inequality in Theorem~\ref{thm:restricted-bl}, only with $\gamma$ and $\beta$ now being functions of $\hat{x}$ and $\hat{y}$ instead of $x^*$ and $y^*$.
%We show there are similar bounds as in the restricted case, though we need to take into account the unknown values of $\hat{x}$ and $\hat{y}$.
The estimate $\hat{w}$ in \eqref{eq:general-bl-opt} is a function of $x^*$, $y^*$, $\hat{x}$, and $\hat{y}$.
Assume we know $\hat{x}$ and $\hat{y}$ as functions of $x^*$ and $y^*$.
We fix bounds on these estimates and show that, in optimizing \eqref{eq:general-bl-opt}, joint estimation either estimates $\hat{w}$ to be one or estimates $\hat{x}$ or $\hat{y}$ to fall outside of specified bounds, indicating a poor estimate in at least one of the three unknown branch parameters.
\begin{restatable}{theorem}{generalBranchInconsist}
\label{thm:general-bl}
Define
\[
\gamma(x, y) = 4xy,
\]
\[
\beta(x, y) = 1+x^2+y^2+x^2y^2,
\]
and bounds
\[
\gamma_L := \gamma_{L}(x^*, y^*) \le \gamma(\hat{x}, \hat{y}),
\]
\[
\gamma_U := \gamma_{U}(x^*, y^*) \ge \gamma(\hat{x}, \hat{y}),
\]
and
\[
\beta_L := \beta_{L}(x^*, y^*) \le \beta(\hat{x}, \hat{y}).
\]
Assume $\hat{x}\neq 0,\hat{y}\neq 0,$ $\hat{x}$ and $\hat{y}$ are not both simultaneously one, and $x^*$ and $y^*$ are not both simultaneously zero.
Then, the maximum likelihood value $\hat{w}=1$ when
\[
-\gamma_{U}^2\left(1 + \frac{1}{2}\left(1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2\right)\right) + 2\gamma_{L}\beta_{L}x^*(y^*)^2 + \beta_{L}^2 \ge 0.
\]
\end{restatable}
We use this theorem to show incorrect branch parameter estimates as follows.
If we do not tolerate any error in pendant branches, we use the tightest possible bounds $\gamma_{L} = \gamma_{U} = \gamma$ and $\beta_{L} = \beta$; this corresponds to the restricted case, analyzed in the Appendix (Fig.~\ref{fig:bl-inconsistency}).
For an intermediate bound, define a specified allowable level of error in estimates for $\hat{x}$ and $\hat{y}$ so that
\[
x^*-x_{L} \le \hat{x} \le x^*+x_{U}
\]
and
\[
y^*-y_{L} \le \hat{y} \le y^*+y_{U}.
\]
Then $\gamma_L, \gamma_U$ and $\beta_L$ can be derived directly from the bounds on $\hat{x}$ and $\hat{y}$.
As an example, Fig.~\ref{fig:bl-loose-inconsistency} shows the case where we have poor estimation of one of the three branch parameters when $x_L=x_U=y_L=y_U=0.1$.
In this case, joint inference will either estimate $\hat{w}$ to have fidelity one or estimate either $\hat{x}$ or $\hat{y}$ to be more than $0.1$ away from its true fidelity.

We provide the following as an intuition for the theoretical development.
For a particular site pattern, in obtaining the joint maximum likelihood function we maximize over ancestral states.
For the internal branch---the branch between the two internal nodes---we have a choice of $(1+w)$ or $(1-w)$ in each of our likelihood terms depending on which ancestral state split corresponds to the highest partial likelihood term.
A maximization procedure will more likely introduce a $(1+w)$ term than a $(1-w)$ term, though this will not be guaranteed as the maximum depends on the values of $x$ and $y$.
This tendency to include $(1+w)$ terms in the likelihood results in a bias toward one of branch fidelities.
If we allow multifurcating trees in our inference, then we can think of this as another instance of converging to the wrong topology.

\begin{figure}
\centering
% obtained by running
% python joint_inf_plot.py --analytic --general-branch-lengths --delta .001 --plot-name figures/bl-loose-inconsistency.svg
\includegraphics[width=\textwidth]{bl-loose-inconsistency-inkscape}
\caption{
    Analytically-derived region of branch parameter inconsistency in terms of probabilities of a character change with the same data-generating setup as Fig.~\ref{fig:inconsistency-farris}.
    In the marked region, either $\hat{p}_w=0$ or one of $p_{x^*}-.05 \le \hat{p}_x \le p_{x^*}+.05$ or $p_{y^*}-.05 \le \hat{p}_y \le p_{y^*}+.05$ will not be true, resulting in poor estimation of the pendant branch parameters.
    In terms of fidelities, these bounds are $x^*-0.1 \le \hat{x} \le x^*+0.1$ and $y^*-0.1 \le \hat{y} \le y^*+0.1$.
    As in previous plots, the shaded and white regions are loose indications of inconsistency.
}
\label{fig:bl-loose-inconsistency}
\end{figure}

@@

Direct numerical optimization confirms our theoretically-derived bounds and provides a more detailed picture compared to the conservative analytically-derived region (Fig.~\ref{fig:bl-loose-inconsistency}).
To determine how conservative, we use the method of basin-hopping \citep{Wales1997} to perform joint estimation (Fig.~\ref{fig:bl-general-inconsistency}).
We see that the region of inconsistency in the general case is similar to that of the restricted case (compare Figs.~\ref{fig:bl-inconsistency} and~\ref{fig:bl-general-inconsistency}).
This region encompasses the majority of the branch parameter space; even given the correct topology and performing our best possible optimization, we have many situations where we will estimate the interior branch length to be zero.

We provide a full description of our optimization procedure in the Appendix, but briefly, we perform two maximizations---one over $0 \le x,y,w \le 1$ and one over $0 \le x,y \le 1$ with $w=1$---and take the value of $\hat{w}$ with the higher objective function.
We compute these maxima over a lattice in steps of $10^{-2}$ for $x^*$ and $y^*$ from $10^{-2}$ to $1-10^{-2}$.
We do not include zero or one in our lattice to further stabilize the fits, as these cases can result in pathologies.
Our optimization code can be found at \url{https://github.com/matsengrp/joint-inf/}.

As data were generated with parameters $\{x^*, y^*, x^*, y^*, y^*\}$, the true value for $w$ is $y^*$.
Marginal inference performs as expected, where $\hat{w}$ is equal to $y^*$ regardless of the value of $x^*$ (Fig.~\ref{fig:bl-general-marginal}) when optimizing \eqref{eq:marginal_likelihood} using the same procedure.
For joint inference, the estimates for $\hat{w}$ when $x^*$ and $y^*$ are both large look reasonable, with $\hat{w}$ increasing as $y^*$ increases, though Fig.~\ref{fig:bl-general-bias} shows there is a systematic positive bias in this procedure even when $\hat{w}$ is not estimated to be one.
To understand the quality of each fit, we report the range of $\hat{w}-y^*$ where $\hat{w}\neq 1$.
For joint inference, the errors range from $[-7\times 10^{-3}, 8\times 10^{-2}]$ and for marginal inference, $[-8\times 10^{-8}, 5\times 10^{-7}]$ showing that, even in cases where joint inference does not estimate $\hat{w}$ to be exactly one, it still fails to achieve a low error from truth.

\begin{figure}
\centering
% obtained by running
% python joint_inf_plot.py --empirical --general-branch-lengths --delta .01 --plot-name figures/w-hat-empirical-01.svg --out-pkl-name figures/w-hat-empirical-01.pkl --n-jobs 16
% or
% python joint_inf_plot.py --empirical --general-branch-lengths --plot-name figures/w-hat-empirical-01.svg --in-pkl-name ../intermed/w-hat-empirical-01.pkl
% if fit already
\includegraphics[width=\textwidth]{w-hat-empirical-01}
\caption{
    Numerical estimates for $\hat{p}_w=(1-\hat{w})/2$ when computing $(\hat{x}, \hat{y}, \hat{w})$ using basin-hopping \citep{Wales1997} optimizing \eqref{eq:profile_likelihood}.
    Data generated as in Fig.~\ref{fig:inconsistency-farris}.
}
\label{fig:bl-general-inconsistency}
\end{figure}

@@

We have shown that jointly inferring ancestral states and phylogenetic parameters \citep{Sagulenko2018-xl} is not consistent in general.
Specifically, in the case of four-taxon trees with infinite data, we have obtained nontrivial regions of generating parameters that result in two types of inconsistency: first, where joint inference converges on the incorrect topology and, second, where it estimates severely biased branch lengths even given the correct topology.
In all cases, these regions of inconsistency arise when the branches of the generating trees are ``long,'' that is, when branch fidelities tend to be small.
This inconsistency in the case of long branches concurs with some empirical findings in \citet{Sagulenko2018-xl}, namely their Figures~2 and 3.
\citet{Sagulenko2018-xl} explicitly warn their approximation is only valid for ``short'' branches, though Fig.~\ref{fig:bl-general-bias} shows even when branches are short there is a consistent bias.

@@

Previous work in phylogenetics has developed consistency counterexamples using the same four-taxon topologies used here \citep{Felsenstein1978-rr}.
In this previous work, when simulating data under the Felsenstein topology $\tau_2$, as the number of observations increases, the InvFels topology $\tau_1$ becomes more likely when performing a particular estimation procedure.
This is the converse of what we have shown for joint inference, where the Felsenstein topology is more likely than the InvFels topology.
Moreover, the inconsistency demonstrated by \citet{Felsenstein1978-rr} is attributed to long branch attraction, i.e., the fact that there may be multiple long branches where parallel changes are more likely than a single change along a short branch.
This is not the case here; for our case, the inconsistency generally occurs when all branches are long, and has more to do with the choices of the form of the likelihood than from the interplay of long and short branches.
Difficulties in phylogenetic estimation when generating data on the InvFels tree have been found by \citet{Siddall1998-hq}, though \citet{Swofford2001-hr} show that sequence length plays a major role in these issues.

The case of joint inference of a phylogenetic likelihood was discussed in \citet{Goldman1990-dk}.
In it, Goldman provides a worked example in which estimating a topology with fixed branch lengths is equivalent to parsimony and thus not guaranteed to be consistent, though he does not discuss the inconsistency of joint inference in general.
We show cases where the estimates of the topology will be inconsistent if we are estimating both topology and branch lengths, and, moreover, even if we know the correct topology, we show cases where branch lengths will also be inconsistent.
Finally, just prior to his conclusion, he discusses when parsimony will give the same answer as maximum likelihood, concluding that the question is ill-posed since parsimony estimates different parameters than maximum likelihood, i.e., it assumes equal branch lengths.
The joint inference procedure outlined here estimates the same parameters as classical maximum likelihood---albeit implicitly estimating ancestral states as well---rendering the question well-posed.
Finally, we are able to provide much more detail on how large branch lengths must be for general joint inference---estimating topology and branch lengths together---to fail to be consistent.

While we have shown inconsistency in both topology and branch parameter estimation, there is substantial scope for future work to make these results more precise and more general.
The techniques used to obtain upper and lower bounds for the likelihoods in the topology estimation case provide relatively loose bounds, though how loose they are remain unknown without either further analysis or verification through simulation.
Similarly, for the general case in estimating branch lengths, we were only able to provide a conservative region of overestimation, and the unusual shape we observe via numerical optimization (Fig.~\ref{fig:bl-general-inconsistency}) begs further investigation.
Empirical validation shows that the general case is not unlike the restricted case.
All of these results hold only for a simple binary symmetric model on four-taxon trees, and extensive simulation is necessary to understand how these results extend to more complicated general cases.
Given that many of the bounds presented here are in the form of level sets of multivariate polynomials, a more formal approach using algebraic geometric techniques may reveal more stable or interesting patterns of inconsistency; see \citet{Sturmfels2002} for a thorough treatment of solving systems of polynomial equations.
Finally, all of the material presented here concerns joint estimation under maximum likelihood, and does not pose any problem for other settings, such as joint sampling of trees and ancestral sequences in a Bayesian framework.

\section*{Acknowledgements}
We thank Richard Neher, Vladimir Minin, and Joe Felsenstein for helpful discussions.

>>>>

\begin{abstract}
Maximum likelihood estimation in phylogenetics requires a means of handling unknown ancestral states.
Classical maximum likelihood averages over these unknown intermediate states, leading to consistent estimation of the topology and continuous model parameters.
Recently, a computationally-efficient approach was proposed to jointly maximize over these unknown states and phylogenetic parameters.
Although this method of joint maximum likelihood estimation can obtain estimates more quickly, its properties as an estimator are not yet clear.
In this paper, we show that this method of jointly estimating phylogenetic parameters along with ancestral states is not consistent in general.
%TODO: change this line? it's still like an inconsistency in topology
We find a set of parameters that generate data under a four-taxon tree for which this joint method estimates an incorrect topology in the limit of infinite-length sequences.
For branch length estimation on the correct topology, we outline similar cases where branch length estimates are consistently and heavily biased.
\end{abstract}

@@

In this paper, we show that the joint inference of trees and ancestral sequences is not consistent in general.
To do so, we use a binary symmetric model with data being generated on the four-taxon ``inverse Felsenstein'' tree (a.k.a.\ the ``Farris'' tree) and we construct bounds on the joint objective function to demarcate a sizeable area of branch lengths in which joint inference is guaranteed to give the wrong tree in the case of perfect sequence data with an infinite number of sites.
We find similar areas where joint inference consistently underestimates interior branch lengths when the topology is known and fixed.

@@

The alternative approach \citep{Sagulenko2018-xl} does away with the marginalization and directly estimates the maximum likelihood parameters of the fully-observed likelihood in \eqref{eq:full_likelihood}.
This is known in statistics as a profile likelihood \citep{Murphy2000-ry}, which exists here because $\alphabet$ is a finite set:
\begin{equation}
\label{eq:profile_likelihood}
L_\nCols'(\tau, t; \fullAlignment) = \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i\in\alphabet^{\nAncestralStateRows}} \ \Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) = \max_{\fullAncestralStates\in\alphabet^{\nAncestralStateRows\times\nCols}} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates).
\end{equation}
We use $\hat{\fullAncestralStates}$ to denote an estimate for $\fullAncestralStates$ obtained by maximizing \eqref{eq:profile_likelihood}, and estimate a topology and branch fidelities using this profile likelihood as
\begin{equation}
\label{eq:profile_likelihood_topology_bl}
(\hat{\tau}, \hat{t}) = \argmax_{\tau, t} \ L_\nCols'(\tau, t; \fullAlignment).
\end{equation}
In general, the functional form of \eqref{eq:profile_likelihood} is determined by inequalities arising from taking maxima over ancestral state splits in Table~\ref{tab:farris_likelihoods} to obtain the likelihood term for each most likely ancestral state, these terms depending on the unknown $(\tau, t)$.
For this reason, in practice, the joint inference strategy estimates $\hat{\fullAncestralStates}$ for a fixed $(\tau,t)$, then $(\hat{\tau},\hat{t})$ given $\hat{\fullAncestralStates}$, maximizing each of these conditional objectives until convergence \citep{Sagulenko2018-xl}.

@@

To show an inconsistency in topology estimation, we start with true generating parameters $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ on the InvFels topology (Fig.~\ref{fig:farris-fels-top}a).
%We show that, as $\nCols\rightarrow\infty$, there exist values for $x^*$ and $y^*$ such that the value of the likelihood after maximizing using joint inference is greater for the Felsenstein topology ($\tau_2$) than for the true, generating InvFels topology ($\tau_1$).
%To do so, we construct an upper bound $C_0(x^*, y^*)$ for the likelihood given the InvFels topology as a function of $x^*$ and $y^*$ and, similarly, a lower bound $C_1(x^*, y^*)$ for the likelihood given the Felsenstein topology.
%When $C_0(x^*, y^*) < C_1(x^*, y^*)$, the likelihood in the Felsenstein case is larger than the likelihood in the InvFels case, demonstrating inconsistency (Fig.~\ref{fig:inconsistency-farris}).

\begin{restatable}{theorem}{topoInconsist}
Let $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ and $t=\{x_1, y_1, x_2, y_2, w\}$.
There exists a set of $0 < x^*, y^* < 1$ such that
\[
\{\hat{x}_1,\hat{y}_1,\hat{x}_2,\hat{y}_2,\hat{w}\} = \arg\max_{t} \ \ell_{\tau_1,t^*}(\tau_1, t)
\]
and $\hat{w}\equiv 1$.
\end{restatable}

In words, this implies an inconsistency as we converge on the incorrect topology, estimating the interior branch length to be zero.
The proof of this theorem is by a detailed examination of inequalities.
%Intuitively, $\tau_2$ is favored in performing joint inference since the objective function for $\tau_2$ has more ``degrees of freedom''---Table~\ref{tab:likelihoods} shows that $\tau_1$ has only three possible forms for its objective function while $\tau_2$ has many more.
%This enables more possible maxima for $\tau_2$ even if data are generated from $\tau_1$.

%TODO: change legend of this figure---the intersection is just the region where \emptyset is maximal, so it will be the same plot
\begin{figure}
\centering
\includegraphics[width=.95\textwidth]{incons-analytic-inkscape}
\caption{
    An analytically-derived region of inconsistency in terms of probabilities of a character change along a branch for ``perfect'' data generated on the InvFels topology (Fig.~\ref{fig:farris-fels-top}) with $p_{w^*} = p_{y^*}$ (in terms of fidelities, $w^*=y^*$).
    Due to the looseness of the upper and lower bounds, the parameters in the white region do not necessarily indicate consistency, though all parameters in the shaded region result in an inconsistency.
}
\label{fig:incons-analytic}
\end{figure}

We provide the following as an intuition for the theoretical development.
For a particular site pattern, in obtaining the joint maximum likelihood function we maximize over ancestral states.
For the internal branch---the branch between the two internal nodes---we have a choice of $(1+w)$ or $(1-w)$ in each of our likelihood terms depending on which ancestral state split corresponds to the highest partial likelihood term.
A maximization procedure will more likely introduce a $(1+w)$ term than a $(1-w)$ term, though this will not be guaranteed as the maximum depends on the values of $x$ and $y$.
This tendency to include $(1+w)$ terms in the likelihood results in a bias toward one of branch fidelities.
This is apparent in the ``long $x^*$, short $y^*$'' scenario as these are the cases in which $\emptyset$ is the most likely ancestral state split (Table~\ref{tab:likelihoods} in the Appendix with $x_1=x_2=x^*$ and $y_1=y_2=y^*$).
If we allow multifurcating trees in our inference, then we can think of this as another instance of converging to the wrong topology.

@@

Direct numerical optimization confirms our theoretically-derived bounds and provides a more detailed picture compared to the conservative analytically-derived region (Fig.~\ref{fig:incons-analytic}).
To determine how conservative, we use the well-known L-BFGS-B method to perform joint estimation (Fig.~\ref{fig:bl-general-inconsistency}).
We see that the region of inconsistency in the general case is much larger than in the analytic case---the analytic case required $\hat{y}_2 = 1$ and dealt with loose bounds, so this is reasonable.
This region encompasses almost half of the branch parameter space; even given the correct topology and performing our best possible optimization, we have many situations where we will estimate the interior branch length to be zero.

We provide a full description of our optimization procedure in the Appendix, but briefly, we perform 81 maximizations for each separate likelihood function.
For each function, we optimize the objective starting at five separate initial values to mitigate problems with small curvature of the objective surface.
We compute these maxima over a lattice in steps of $10^{-2}$ for $x^*$ and $y^*$ from $10^{-2}$ to $1-10^{-2}$.
We do not include zero or one in our lattice to further stabilize the fits, as these cases can result in pathologies.
Our optimization code can be found at \url{https://github.com/matsengrp/joint-inf/}.

As data were generated with parameters $\{x^*, y^*, x^*, y^*, y^*\}$, the true value for $w$ is $y^*$.
Marginal inference performs as expected, where $\hat{w}$ is equal to $y^*$ regardless of the value of $x^*$ (Fig.~\ref{fig:bl-general-marginal}) when optimizing \eqref{eq:marginal_likelihood} using the same procedure.
For joint inference, the estimates for $\hat{p}_w$ when $p_{x^*}$ and $p_{y^*}$ are both small look reasonable, with $\hat{p}_w$ increasing as $p_{y^*}$ increases, though Fig.~\ref{fig:empirical-bias} shows there is a systematic negative bias in this procedure even when the true branches are short.
To understand the quality of each fit, we report the range of $\hat{p}_w-p_{y^*}$ where $p_{x^*}, p_{y^*} \le .1$.
Due to the discontinuities in the fit (Fig.~\ref{fig:bl-general-inconsistency}), an overall estimate of the bias is misleading, and these short-branch cases should be the best settings for joint optimization.
For joint inference, the errors range from $[-2\times 10^{-2}, 3\times 10^{-3}]$ and for marginal inference, $[-3\times 10^{-17}, 3\times 10^{-17}]$ showing that, even in cases where joint inference is supposed to do well, it still fails to achieve a low error from truth.

\begin{figure}
\centering
\includegraphics[width=\textwidth]{empirical-estimate}
\caption{
    Numerical estimates for $\hat{p}_w=(1-\hat{w})/2$ when computing $(\hat{x}, \hat{y}, \hat{w})$ optimizing \eqref{eq:profile_likelihood}.
    Data generated as in Fig.~\ref{fig:incons-analytic}
}
\label{fig:bl-general-inconsistency}
\end{figure}

@@

We have shown that jointly inferring ancestral states and phylogenetic parameters \citep{Sagulenko2018-xl} is not consistent in general.
Specifically, in the case of four-taxon trees with infinite data, we have obtained nontrivial regions of generating parameters that result in a type of inconsistency: the joint inference procedure estimates severely biased branch lengths even given the correct topology.
These regions of inconsistency arise when the top two branches of the generating trees are ``long,'' that is, when the top branch fidelities tend to be small, and when the lower branches are ``short,'' i.e., have large fidelities.
This inconsistency in the case of long branches concurs with some empirical findings in \citet{Sagulenko2018-xl}, namely their Figures~2 and 3.
\citet{Sagulenko2018-xl} explicitly warn their approximation is only valid for ``short'' branches, though Fig.~\ref{fig:empirical-bias} shows even when branches are short there is a consistent bias.

@@

Previous work in phylogenetics has developed consistency counterexamples using the same four-taxon topologies used here \citep{Felsenstein1978-rr}.
In this previous work, when simulating data under the Felsenstein topology $\tau_2$, as the number of observations increases, the InvFels topology $\tau_1$ becomes more likely when performing a particular estimation procedure.
We have shown cases in which, when generating from the InvFels topology, we converge to a multifurcating topology, with one or more branch lengths estimated to be zero.
Moreover, the inconsistency demonstrated by \citet{Felsenstein1978-rr} is attributed to long branch attraction, i.e., the fact that there may be multiple long branches where parallel changes are more likely than a single change along a short branch.
This is not the case here; while analytically the inconsistency occurs in the long branch attraction region of the parameter space, we see empirically that this inconsistency is present in roughly half of the entire parameter space, and occurs when the true branches generate data that more likely has no change along the interior branch.
Difficulties in phylogenetic estimation when generating data on the InvFels tree have been found by \citet{Siddall1998-hq}, though \citet{Swofford2001-hr} show that sequence length plays a major role in these issues.

The case of joint inference of a phylogenetic likelihood was discussed in \citet{Goldman1990-dk}.
In it, Goldman provides a worked example in which estimating a topology with fixed branch lengths is equivalent to parsimony and thus not guaranteed to be consistent, though he does not discuss the inconsistency of joint inference in general.
We show cases where the estimates of the topology are inconsistent if we are estimating both topology and branch lengths, and, moreover, even if we know the correct topology, we show cases where branch lengths are severely biased.
Finally, just prior to his conclusion, he discusses when parsimony gives the same answer as maximum likelihood, concluding that the question is ill-posed since parsimony estimates different parameters than maximum likelihood, i.e., it assumes equal branch lengths.
The joint inference procedure outlined here estimates the same parameters as classical maximum likelihood---albeit implicitly estimating ancestral states as well---rendering the question well-posed.
Finally, we are able to provide much more detail on how large branch lengths must be for general joint inference---estimating topology and branch lengths together---to fail to be consistent.

While we have shown inconsistency in both topology and branch parameter estimation, there is substantial scope for future work to make these results more precise and more general.
The techniques used to obtain upper and lower bounds for the likelihoods in the topology estimation case provide relatively loose bounds, though how loose they are remain unknown without either further analysis or verification through simulation.
Similarly, for the general case in estimating branch lengths, we were only able to provide a conservative region of overestimation, and the discontinuities we observe via numerical optimization (Fig.~\ref{fig:bl-general-inconsistency}) begs further investigation.
Empirical validation shows that the region of inconsistency is much larger than the one analytically derived.
All of these results hold only for a simple binary symmetric model on four-taxon trees, and extensive simulation is necessary to understand how these results extend to more complicated general cases.
Given that many of the bounds presented here are in the form of level sets of multivariate polynomials, a more formal approach using algebraic geometric techniques may reveal more stable or interesting patterns of inconsistency; see \citet{Sturmfels2002} for a thorough treatment of solving systems of polynomial equations.
Finally, all of the material presented here concerns joint estimation under maximum likelihood, and does not pose any problem for other settings, such as joint sampling of trees and ancestral sequences in a Bayesian framework.

\section*{Acknowledgements}
We thank Richard Neher, Vladimir Minin, and Joe Felsenstein for helpful discussions.
Insight from the editors and Vu Dinh greatly improved this manuscript.
